---
title: "IFNy_DEG_Analysis"
author: "Harsha Tamtam"
date: "2026-06-05"
output: html_document
---
```{r setup, include=FALSE}
#this chuck is to setup defaults for basic settings of the webpage, note that this chunk is R language and the chunk below is CSS language
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	dev = c("png","pdf")
)
```

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

## Library Section
```{r Libraries}
library(tidyverse)
library(dplyr)
library(tibble)
library(scales)
library(edgeR)
library(smplot2)
library(ggpubr)
library(workflowr)
library(ComplexHeatmap)
library(RColorBrewer)
library(gridExtra)
library(ggrepel)
library(ggVennDiagram)
library(grid)
library(ggrepel)
library(org.Hs.eg.db)
library(enrichplot)
library(clusterProfiler)
library(circlize)
library(gprofiler2)
library(stringr)
```

## File Calling

Reading in files:
```{r}
metadata <- readRDS("data/metadata_IFNy.RDS")
counts <- readRDS("data/counts_IFNy.RDS")

raw_counts <- counts %>% 
  dplyr::select(-c(Entrez_ID, Symbol)) %>% 
  as.data.frame() %>% 
  rename(., Ensembl = Ensemble) %>% 
  pivot_longer(-Ensembl, names_to = "Samples", values_to = "raw_counts") %>% 
  mutate(raw_counts = as.numeric(raw_counts)) %>% 
  left_join(metadata, by = c("Samples" = "SampleName")) %>% 
  select(Ensembl, raw_counts, Condition) %>% 
  pivot_wider(names_from = Condition, values_from = raw_counts) %>% 
  column_to_rownames("Ensembl")


```

Converting to Log2 CPM and filtering rowMeans log2cpm > 0:
```{r}
raw_log2cpm <- cpm(raw_counts, log = TRUE)
filtered_raw_counts <- raw_counts[rowMeans(raw_log2cpm)> 0,]

# dim(filtered_raw_counts)
# str(filtered_raw_counts)
```


## Identifying DGEs
```{r}
metadata2 <- metadata %>%
  filter(!Condition %in% c("C_IFNy_C2R", "C_VEH_C2R"))

filtered_counts2 <- filtered_raw_counts %>%
  dplyr::select(-c(C_IFNy_C2R, C_VEH_C2R))

# Create DGEList
dge <- DGEList(
  counts = filtered_counts2
)
# TMM normalization
dge <- calcNormFactors(
  dge,
  method = "TMM"
)
# Ensuring my factors are set correctly
metadata2$Species <- relevel(
  factor(metadata2$Species),
  ref = "C")
metadata2$Treatment <- relevel(
  factor(metadata2$Treatment),
  ref = "VEH")
# The code below includes C2R replicate is being considered for normalization, but we are xcluding it, hence why this code is cut out
# metadata$Cond2 <- metadata$Cond
# metadata$Cond2 <- gsub("_R$", "", metadata$Cond2)

# in the 'design' object, we would include '+ Cond2' if we are considering C2R replicate in the following calculations
design <- model.matrix(~ Species * Treatment,
  data = metadata2
)
#colnames(design) <- gsub("^Cond2", "", colnames(design))
colnames(design) <- c(
  "Intercept",
  "SpeciesH",
  "TreatmentIFNy",
  "SpeciesH_TreatmentIFNy"
)

cor_fit <- duplicateCorrelation(object = dge$counts,
                                design = design,
                                block = metadata2$Individual
                                )

voom_trend <- voom(dge,
                   design,
                   block = metadata2$Individual, 
                   correlation = cor_fit$consensus.correlation,
                   plot = FALSE)
```

```{r Chunk of code observing number of DEGs across treatment}
# Fit linear model
fit <- lmFit(
  voom_trend,
  design,
  block = metadata2$Individual,
  correlation = cor_fit$consensus.correlation
)
# Defining contrasts
contrast_matrix <- makeContrasts(
  Human_vs_Chimp_VEH = SpeciesH,
  Human_vs_Chimp_IFNy = SpeciesH + SpeciesH_TreatmentIFNy,
  Human_IFNy_vs_VEH = TreatmentIFNy + SpeciesH_TreatmentIFNy,
  Chimp_IFNy_vs_VEH = TreatmentIFNy,
  Interaction = SpeciesH_TreatmentIFNy,
  levels = design)
# Applying contrasts across samples
fit2 <- contrasts.fit(fit, contrast_matrix)
efit2 <- eBayes(fit2)
signif_gene_results = decideTests(efit2, adjust.method = "BH", p.value = 0.05)

#summary(signif_gene_results)
#plotSA(efit2, main = "Mean-variance final model trend")

```
## Top Tables

```{r Creating TopTables for Plots}

V.Human_vs_Chimp_VEH <- topTable(efit2, coef = "Human_vs_Chimp_VEH", adjust.method = "BH", p.value = 1, number = Inf)
V.Human_vs_Chimp_IFNy <- topTable(efit2, coef = "Human_vs_Chimp_IFNy", adjust.method = "BH", p.value = 1, number = Inf)
V.Human_IFNy_vs_VEH <- topTable(efit2, coef = "Human_IFNy_vs_VEH", adjust.method = "BH", p.value = 1, number = Inf)
V.Chimp_IFNy_vs_VEH <- topTable(efit2, coef = "Chimp_IFNy_vs_VEH", adjust.method = "BH", p.value = 1, number = Inf)
V.Interaction <- topTable(efit2, coef = "Interaction", adjust.method = "BH", p.value = 1, number = Inf)

save_toptable_list <- list("HvC_VEH"=V.Human_vs_Chimp_VEH, "HvC_INFy"=V.Human_vs_Chimp_IFNy, "IFNyvVEH_H"=V.Human_IFNy_vs_VEH, "IFNyvVEH_C"=V.Chimp_IFNy_vs_VEH, "Interaction"=V.Interaction)

saveRDS(save_toptable_list,"data/toptable_results.RDS")
```

```{r Interaction Venn Diagrams}

all_genes <- save_toptable_list %>%
  purrr::imap(~ .x %>%
                tibble::rownames_to_column("Ensembl") %>%
                mutate(source = .y)) %>%
  bind_rows()

sig_results <- all_genes %>%
  filter(adj.P.Val < 0.05)

sig_venn_list <- split(
  sig_results$Ensembl,
  sig_results$source
)
 
sig_venn_treatments <- sig_venn_list[c("IFNyvVEH_H", "IFNyvVEH_C")]
sig_venn_species <- sig_venn_list[c("HvC_VEH", "HvC_INFy")]

ggVennDiagram::ggVennDiagram(
  sig_venn_treatments,
  label_alpha = 0,
) +
  scale_fill_gradient(
    low = "white",
    high = "maroon",
    limits = c(1000, 2500)
  ) +
  theme_void()+
  xlim(-5, 5)+
  coord_flip()

ggVennDiagram::ggVennDiagram(
  sig_venn_species,
  label_alpha = 0,
) +
  scale_fill_gradient(
    low = "white",
    high = "maroon",
    limits = c(0, 6000)
  ) +
  theme_void() +
  xlim(-5, 5)+
  coord_flip()

```
## Volcano Plots
```{r Volcano Plots comparing treatment conditions within the same species using limma, fig.height=4, fig.width=6}
toptable_list <- readRDS("data/toptable_results.RDS")
#toptable_list <- list(V.Human_IFNy_vs_VEH, V.Chimp_IFNy_vs_VEH, V.IFNy_Human_vs_Chimp, V.VEH_Human_vs_Chimp)
comparison_name_list <- list("VEH: Human vs Chimp", "IFNy: Human vs Chimp", "Human: IFNy vs VEH", "Chimp: IFNy vs VEH", "Interaction")

generate_volcano_plot <- function(toptable, comparison_name) {

  volcano_colors <- toptable %>%
    rownames_to_column("Genes") %>%
    left_join(counts %>% 
                select(Ensemble, Symbol),
              by = c("Genes" = "Ensemble")) %>% 
    mutate(
    id_label = ifelse(
      is.na(Symbol),
      Genes,
      Symbol
    )
  ) %>% 
    mutate(
      neglog10AdjP = -log10(pmax(adj.P.Val, 1e-300)),
      color_key = case_when(
        adj.P.Val < 0.05 & logFC > 0  ~ "Up",
        adj.P.Val < 0.05 & logFC < 0 ~ "Down",
        TRUE ~ "NS"
      )
    )

  top_genes <- volcano_colors %>%
  arrange(adj.P.Val) %>%
  slice_head(n = 10)
  
  # genes_of_interest <- volcano_colors %>%
  #   dplyr::filter(id_label %in% c("ASPHD2", "C1R", "PARP12", "GBP2", "TNFSF13B", "TUBB", "GRAMD2A", "FAM107A", "ZCWPW2", "TRIB1", "C1S", "CD47", "SFT2D2", "CIBAR1", "IFITM2", "TESK2", "STAMBPL", "NUP58", "BPGM", "CIMAP3", "CD9", "CFH", "SLC35A4", "ENSG00000284048", "UST", "A2M", "CACHD1", "VPS13C", "FBXO31", "NORAD"))

  p <- ggplot(
    volcano_colors,
    aes(x = logFC, y = neglog10AdjP)
    ) +
    geom_point(
      aes(color = color_key),
      size = 2,
      alpha = 0.6
    ) +
    
   # coord_cartesian(
   #    xlim = c(-12, 12),
   #    ylim = c(0, 20)
   #  ) +
    
    # geom_vline(
    # xintercept = c(-1,1),
    # linetype = "dashed"
    # ) +
    # geom_hline(
    # yintercept = -log10(0.05),
    # linetype = "dashed"
    # ) +
    
    geom_text_repel(
      data = top_genes,
      #data = genes_of_interest,
      aes(label = id_label),
      size = 3,
      max.overlaps = Inf,
      box.padding = 0.5
    ) +
    
    scale_color_manual(
      values = c(
        "Up" = "red",
        "Down" = "blue",
        "NS" = "grey50"
      )
    ) +
    labs(
      title = comparison_name,
      x = "log2FC",
      y = "-log10(Adj.P-value)"
    ) +
    theme_classic() +
    
    theme(
      plot.title = element_text(hjust = 0.6),
      text = element_text(size = 12)
    ) 
  
    list(
      plot = p,
      top = top_genes
      #interesting_genes = genes_of_interest
      )
}

for (i in seq_along(toptable_list)) {
  #Generate and display the results
  
  results <- generate_volcano_plot(
    toptable_list[[i]],
    comparison_name_list[[i]]
  )
print(results$top)
#print(results$interesting_gene)
print(results$plot) # Print the plot to view
}
#interesting genes: GRAMD2B, ARG2, maybe PITPNM2, CD274 (encodes PD-L1), GBP4
```
## Boxplots

```{r Preparing log2 cpm counts and binding counts to show Human and Chimp designations}
log2_filtered_counts <-
  filtered_counts2 %>% 
  cpm(., log=TRUE)%>%
  as.data.frame() 

row.names(log2_filtered_counts) <- row.names(filtered_counts2)

human_deg <- V.Human_IFNy_vs_VEH %>%
  tibble::rownames_to_column("Ensembl") %>%
  mutate(species = "Human")
chimp_deg <- V.Chimp_IFNy_vs_VEH %>%
  tibble::rownames_to_column("Ensembl") %>%
  mutate(species = "Chimp")

deg_species <- bind_rows(human_deg, chimp_deg)
```

```{r Boxplots showing top most significant DEGs filtering to consider p less than 0.05 between Species and Treatments, fig.width=5, fig.height=5}
top_genes <- deg_species %>%
  group_by(species) %>%
  #Including the above line will give top 'n' genes for each species. Note that even with stating 'n = number', not all genes may be plotted due to duplicates. Exclude to only show top 'n' across both species
  
  # arrange(adj.P.Val) %>% 
  # slice_head(n=8) %>% 
  dplyr::filter(adj.P.Val < 0.05) %>% 
  slice_min(adj.P.Val, n = 6) %>% 
  #The above line can be replaced with 'arrange(adj.P.Val) %>% slice_head(n-8) %>%'. Remember, slice_head:slice_min::slice_tail::slice_max
  pull(Ensembl)
#head(top_genes)


log2_filtered_df <- log2_filtered_counts %>% 
  tibble::rownames_to_column("Ensemble") %>%
  left_join(counts %>% select(Ensemble, Symbol),
            by = "Ensemble") %>%
    mutate(id_label = ifelse(
      is.na(Symbol),
      Ensemble,
      Symbol)
      ) %>%
  dplyr::filter(Ensemble %in% top_genes) %>%
  #log2_filtered_df
  pivot_longer(cols = starts_with("C_") | starts_with("H_"), names_to = "sample", values_to = "log2_counts") %>% 
      # We can use this code as an alternative -> pivot_longer(cols = !c(Ensemble, Symbol, id_label), names_to = "sample", values_to = "log2_counts") %>%
      # This is another alternative -> pivot_longer(cols = matches("^[CH]_"), names_to = "sample", values_to = "log2_counts")
  separate("sample", into = c("species","trt","ind")) %>%
      #mutate(spec_trt_group = paste(spec, trt, sep = "_")) %>% 
  mutate(species=factor(species, levels = c("C","H"), labels = c("Chimp", "Human"))) %>% 
  mutate(trt=factor(trt, levels= c("VEH","IFNy")))

ggplot(log2_filtered_df,
       aes(x = trt, y = log2_counts, fill = trt)) +
  geom_boxplot(aes(fill = trt)) +
  geom_point(aes(color = ind)) +
  facet_grid(species ~ id_label) +
  scale_fill_manual(
    name = "Treatment",
    values = c("VEH" = "#41B333", "IFNy" = "#8B006D")
  ) +
  labs(
    title = "Top Significant DEGs in Humans and Chimps (may exclude overlapping genes)",
    color = "Individual",
    x = "Treatment",
    y = "log2 CPM"
  ) +
  theme_bw()
```

```{r Boxplots showing bottom most significant DEGs filtering to consider p less than 0.05 between Species and Treatments, fig.width=5, fig.height=5}
bottom_genes <- deg_species %>%
  group_by(species) %>%
  #Including the above line will give top 'n' genes for each species. Note that even with stating 'n = number', not all genes may be plotted due to duplicates. Exclude to only show top 'n' across both species
  
  # arrange(adj.P.Val) %>% 
  # slice_head(n=8) %>% 
  dplyr::filter(adj.P.Val < 0.05) %>% 
  slice_max(adj.P.Val, n = 6) %>% 
  #bottom_genes
  #The above line can be replaced with 'arrange(adj.P.Val) %>% slice_head(n-8) %>%'. Remember, slice_head:slice_min::slice_tail::slice_max
  pull(Ensembl)
#head(bottom_genes)

log2_filtered_df <- log2_filtered_counts %>% 
  tibble::rownames_to_column("Ensemble") %>%
  left_join(counts %>% select(Ensemble, Symbol),
            by = "Ensemble") %>%
    mutate(id_label = ifelse(
      is.na(Symbol),
      Ensemble,
      Symbol)
      ) %>%
  dplyr::filter(Ensemble %in% bottom_genes) %>%
  #log2_filtered_df
  pivot_longer(cols = starts_with("C_") | starts_with("H_"), names_to = "sample", values_to = "log2_counts") %>% 
      # We can use this code as an alternative -> pivot_longer(cols = !c(Ensemble, Symbol, id_label), names_to = "sample", values_to = "log2_counts") %>%
      # This is another alternative -> pivot_longer(cols = matches("^[CH]_"), names_to = "sample", values_to = "log2_counts")
  separate("sample", into = c("species","trt","ind")) %>%
      #mutate(spec_trt_group = paste(spec, trt, sep = "_")) %>% 
  mutate(species=factor(species, levels = c("C","H"), labels = c("Chimp", "Human"))) %>% 
  mutate(trt=factor(trt, levels= c("VEH","IFNy")))

ggplot(log2_filtered_df,
       aes(x = trt, y = log2_counts, fill = trt)) +
  geom_boxplot(aes(fill = trt)) +
  geom_point(aes(color = ind)) +
  facet_grid(species ~ id_label) +
  scale_fill_manual(
    name = "Treatment",
    values = c("VEH" = "#41B333", "IFNy" = "#8B006D")
  ) +
  labs(
    title = "Bottom significant DEGs in Humans and Chimps (may exclude overlapping genes)",
    color = "Individual",
    x = "Treatment",
    y = "log2 CPM"
  ) +
  theme_bw()

```


```{r This code is looking at Species:Treatment log2cpm changes in the most significantly altered gene expressions, fig.height=5, fig.width=5, eval=FALSE, include=FALSE}

log2_filtered_df <- log2_filtered_counts %>% 
  tibble::rownames_to_column("Ensemble") %>%
  left_join(counts %>% select(Ensemble, Symbol),
            by = "Ensemble") %>%
    mutate(id_label = ifelse(
      is.na(Symbol),
      Ensemble,
      Symbol)
      ) %>%
  dplyr::filter(id_label %in% c("NORAD", "ENSG000000284048", "UST", "FBXO31", "CACHD1", "A2M", "VPS13C", "C1R", "ASPHD2", "TESK2", "TUBB", "GBP2", "CFH")) %>%
  mutate(id_label=factor(id_label, levels = c("NORAD", "ENSG000000284048", "UST", "FBXO31", "CACHD1", "A2M", "VPS13C", "C1R", "ASPHD2", "TESK2", "TUBB", "GBP2", "CFH"))) %>% 
  #log2_filtered_df
  pivot_longer(cols = starts_with("C_") | starts_with("H_"), names_to = "sample", values_to = "log2_counts") %>% 
      # We can use this code as an alternative -> pivot_longer(cols = !c(Ensemble, Symbol, id_label), names_to = "sample", values_to = "log2_counts") %>%
      # This is another alternative -> pivot_longer(cols = matches("^[CH]_"), names_to = "sample", values_to = "log2_counts")
  separate("sample", into = c("species","trt","ind")) %>%
      #mutate(spec_trt_group = paste(spec, trt, sep = "_")) %>% 
  mutate(species=factor(species, levels = c("C","H"), labels = c("Chimp", "Human"))) %>% 
  mutate(trt=factor(trt, levels= c("VEH","IFNy"))) 

# Code for a faceted wrap
# ggplot(log2_filtered_df,
#        aes(x = trt, y = log2_counts, fill = trt)) +
#   geom_boxplot(aes(fill = trt)) +
#   geom_point(aes(color = ind)) +
#   facet_grid(id_label ~ species) +
#   scale_fill_manual(
#     name = "Treatment",
#     values = c("VEH" = "#41B333", "IFNy" = "#8B006D")
#   ) +
#   labs(
#     title = "Interaction DEGs in Humans and Chimps",
#     color = "Individual",
#     x = "Treatment",
#     y = "log2 CPM"
#   ) +
#   theme_bw()

# ggplot(
#   log2_filtered_df,
#   aes(
#     x = species,
#     y = log2_counts,
#     fill = trt
#   )
# ) +
#   geom_boxplot(
#     position = position_dodge(width = 0.8)
#   ) +
#   geom_point(
#   aes(
#     color = ind,
#     group = trt
#   ),
#   position = position_dodge(width = 0.8),
#   size = 2
# ) +
#   facet_wrap(~ id_label, scales = "free_y") +
#   scale_fill_manual(
#     values = c(
#       "VEH" = "#41B333",
#       "IFNy" = "#8B006D"
#     )
#   ) +
#   theme_bw()

gene_list <- unique(log2_filtered_df$id_label)

for (gene in gene_list) {

  gene_df <- log2_filtered_df %>%
    filter(id_label == gene)

  p <- ggplot(
    gene_df,
    aes(
      x = species,
      y = log2_counts,
      fill = trt
    )
  ) +
    geom_boxplot(
      aes(group = interaction(species, trt)),
      position = position_dodge(width = 0.8)
    ) +
    geom_point(
      aes(
        color = ind,
        group = trt
      ),
      position = position_dodge(width = 0.8),
      size = 2
    ) +
    scale_fill_manual(
      values = c(
        "VEH" = "#41B333",
        "IFNy" = "#8B006D"
      )
    ) +
    labs(
      title = gene,
      x = NULL,
      y = "log2 CPM"
    ) +
    theme_bw()

  print(p)
}
```

## Human IFNyvVEH vs Chimp IFNyvVEH Gene Expression Correlation

```{r Scatter plot, fig.width=6, fig.height=4}

# colnames(V.Human_IFNy_vs_VEH)
# colnames(counts)
# head(counts)
human_fc <- V.Human_IFNy_vs_VEH %>%
  tibble::rownames_to_column("Ensembl") %>%
  dplyr::select(Ensembl, logFC) %>%
  rename(Human_logFC = logFC)

chimp_fc <- V.Chimp_IFNy_vs_VEH %>%
  tibble::rownames_to_column("Ensembl") %>%
  dplyr::select(Ensembl, logFC) %>%
  rename(Chimp_logFC = logFC)

fc_df <- inner_join(human_fc, chimp_fc, by = "Ensembl") %>%
  left_join(
    counts %>% 
      dplyr::select(Ensemble, Symbol),
    by = c("Ensembl" = "Ensemble")
  )

interaction_sig_genes <- V.Interaction %>%
  tibble::rownames_to_column("Ensembl") %>%
  filter(adj.P.Val < 0.05) %>% 
  pull(Ensembl)

# fc_df_plot <- fc_df %>%
#   mutate(
#     interaction_sig = ifelse(
#       Ensembl %in% interaction_sig_genes,
#       "Significant Interaction",
#       "Other"
#     )
#   )

# label_df <- fc_df_with_interaction %>%
#   filter(Symbol %in% c("ASPHD2", "VPS13C", "A2M", "C1R", "GBP2", "FBXO31", "NORAD"))
# 
# OR
# 
# label_df <- V.Interaction %>%
#   tibble::rownames_to_column("Ensembl") %>%
#   filter(adj.P.Val < 0.05) %>%
#   arrange(adj.P.Val) %>%
#   slice_head(n = 20) %>%
#   inner_join(fc_df, by = "Ensembl")

# ggplot(
#   fc_df_plot,
#   aes(
#     x = Human_logFC,
#     y = Chimp_logFC,
#     color = interaction_sig
#   )
# ) +
#   geom_point(alpha = 0.6) +
# 
#   scale_color_manual(
#     values = c(
#       "Other" = "grey70",
#       "Significant Interaction" = "pink"
#     )
#   ) +


top_interaction_genes <- V.Interaction %>%
  tibble::rownames_to_column("Ensembl") %>%
  filter(adj.P.Val < 0.05) %>%
  slice_head(n = 20) %>% 
  pull(Ensembl)

ggplot(fc_df, aes(Human_logFC, Chimp_logFC)) +

  geom_point(
    color = "grey70",
    alpha = 0.3,
    size = 0.75
  ) +

  geom_point(
    data = fc_df %>%
      filter(Ensembl %in% interaction_sig_genes),
    color = "pink",
    size = 0.75
  ) +
  
  # geom_text_repel(
  #     data = fc_df %>% 
  #       filter(Ensembl %in% top_interaction_genes),
  #     aes(label = Symbol),
  #     size = 3,
  #     max.overlaps = Inf,
  #     box.padding = 0.5
  #   ) +

# geom_text_repel(
#   data = subset(label_df, Symbol != c("A2M", "FBXO31")),
#   aes(label = Symbol),
#   color = "black",
#   size = 4,
#   box.padding = 1.5,
#   point.padding = 0.1,
#   force = 5,
#   segment.color = "black",
#   segment.size = 0.4,
#   max.overlaps = Inf
#   ) +

  coord_cartesian(
  xlim = c(-12, 12),
  ylim = c(-12, 12)
  ) +
  scale_x_continuous(
  breaks = seq(-12, 12, by = 6)
  ) +
scale_y_continuous(
  breaks = seq(-12, 12, by = 6)
  ) +
    geom_vline(
    xintercept = 0,
    linetype = "dashed"
    ) +
    geom_hline(
    yintercept = 0,
    linetype = "dashed"
    ) +
  
  stat_cor(method = "pearson",
  label.x = -11,
  label.y = 11
  ) +
  
  theme_classic() +
  labs(
    x = "Human IFNy vs VEH logFC",
    y = "Chimp IFNy vs VEH logFC",
    title = "Cross-species IFNy response correlation"
  )

# levels(metadata2$Treatment)
# levels(metadata2$Species)
# levels(metadata$Treatment)
# colnames(design)
# 
# colnames(contrast_matrix)
# 
# head(V.Human_IFNy_vs_VEH$logFC)
# head(V.Human_VEH_vs_IFNy$logFC)

```
## GO/KEGG

### gProfiler2 GO:BP Attempt
```{r gProfiler2 BP Human analyses, fig.height=8, fig.width=10}
chimp_bg <- V.Chimp_IFNy_vs_VEH %>%
  tibble::rownames_to_column("Ensembl") %>%
  pull(Ensembl)

chimp_sig <- chimp_deg %>%
  filter(adj.P.Val < 0.05) %>%
  pull(Ensembl)

gost_chimp <- gost(
  query = chimp_sig,
  organism = "hsapiens",
  custom_bg = chimp_bg,
  sources = "GO:BP"
)

human_bg <- V.Human_IFNy_vs_VEH %>%
  tibble::rownames_to_column("Ensembl") %>%
  pull(Ensembl)

human_sig <- human_deg %>%
  filter(adj.P.Val < 0.05) %>%
  pull(Ensembl)

gost_human <- gost(
  query = human_sig,
  organism = "hsapiens",
  custom_bg = human_bg,
  sources = "GO:BP"
)

# chimp_GO <- gost_chimp$result
# chimp_GO

human_GO <- gost_human$result
# human_GO

# gostplot(gost_human)

human_GO %>%
  slice_min(p_value, n = 20) %>%
  mutate(
    term_name = factor(term_name,
                       levels = term_name)
  ) %>%
  ggplot(aes(x = -log10(p_value),
             y = rev(term_name),
             size = intersection_size,
             color = p_value)) +
  geom_point() +
  theme_classic() +
  labs(
    x = "-log10(adjusted p-value)",
    y = NULL,
    size = "GeneCount"
  )
#length(unique(human_GO$term_id))
```


### gProfiler2 KEGG Attempt for Human IFNy vs VEH
```{r KEGG Human analyses, eval=FALSE, include=FALSE}
chimp_sig <- chimp_deg %>%
  filter(adj.P.Val < 0.05) %>%
  pull(Ensembl)

kegg_results_chimp <- gost(
    query = chimp_sig,
    organism = "ptroglodytes",
    sources = "KEGG"
)

human_sig <- human_deg %>%
  filter(adj.P.Val < 0.05) %>%
  pull(Ensembl)

kegg_results_human <- gost(
    query = human_sig,
    organism = "hsapiens",
    sources = "KEGG"
)

kegg_results_human$result
# kegg_results_chimp$result

kegg_human_df <- kegg_results_human$result %>%
    arrange(desc(-log10(p_value))) %>%
    mutate(adj.log.pval = -log10(p_value))


top_kegg_human <- kegg_human_df %>%
    slice_head(n = 20)

# Code to show KEGG results on bar chart
# ggplot(top_kegg_human,
#        aes(x = reorder(term_name, adj.log.pval),
#            y = adj.log.pval)) +
#     geom_col(fill = "steelblue") +
#     coord_flip() +
#     labs(
#         x = NULL,
#         y = "-log10(adjusted p-value)",
#         title = "KEGG Pathway Enrichment for Human: IFNy v VEH"
#     ) +
#     theme_classic()

# Code to show KEGG results on dot plot
ggplot(top_kegg_human,
       aes(x = adj.log.pval,
           y = reorder(term_name, adj.log.pval),
           size = intersection_size,
           color = p_value)) +
    geom_point() +
    labs(
        x = "-log10(adjusted p-value)",
        y = NULL,
        size = "GeneCount",
        title = "KEGG Pathway Enrichment for Human: IFNy v VEH"
    ) +
    theme_classic()
```

### clusterProfiler:enrichGO GO:BP Attempt
```{r enrichGO BP Human analyses, include=FALSE}
human_sig_entrez <- bitr(human_sig,
    fromType = "ENSEMBL",
    toType = "ENTREZID", 
    OrgDb = org.Hs.eg.db)

human_bg_entrez <- bitr(human_bg,
    fromType = "ENSEMBL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db) 

ego_human <- enrichGO(
  gene = human_sig_entrez$ENTREZID,
  universe = human_bg_entrez$ENTREZID,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID", 
  ont = "BP", 
  pAdjustMethod = "BH", 
  readable = TRUE )

#summary(ego_human)

# # Default dot plot from enrichGO
# dotplot(
#   ego_human,
#   showCategory = 20,
#   font.size = 10)

# # ggplot barplot code for enrichGO
# ego_human@result %>%
#   arrange(p.adjust) %>%
#   slice_head(n = 20) %>%
#   mutate(
#     Description = factor(
#       Description,
#       levels = rev(Description)
#     )
#   ) %>%
#   ggplot(aes(
#     x = -log10(p.adjust),
#     y = Description
#   )) +
#   geom_col(fill = "steelblue") +
#   theme_classic() +
#   labs(
#     x = expression(-log[10]("Adjusted p-value")),
#     y = NULL,
#     title = "Human GO Biological Processes"
#   )

ego_human@result %>%
  arrange(p.adjust) %>%
  slice_head(n = 20) %>%
  mutate(
    Description = factor(
      Description,
      levels = rev(Description)
    )
  ) %>%
  ggplot(aes(
    x = -log10(p.adjust),
    y = Description,
    size = Count,
    color = p.adjust
  )) +
  geom_point() +
  scale_color_gradient(low = "red", high = "blue") +
  theme_classic() +
  labs(
    x = expression(-log[10]("adjusted p-value")),
    y = NULL
  )

```

### Venn Diagram depicting unique and shared Human: IFNy v VEH GO_BP_terms from gprofiler and enrichGO
```{r Venn Diagram reporting number of unique and shared Human IFNy v VEH GO_BP_terms identified from enrichGO and gProfiler2}

enrichGO_terms <- ego_human@result %>%
  filter(p.adjust < 0.05) %>%
  pull(ID)

gprofiler2_terms <- gost_human$result %>%
  filter(p_value < 0.05) %>%
  pull(term_id)

ggVennDiagram(
  list(
    clusterProfiler = enrichGO_terms,
    gProfiler2 = gprofiler2_terms
  )
) +
scale_fill_gradient(low = "light blue", high = "dark blue") +
theme_void() +
#ggtitle("Number of unique GO:BP terms identified from GO codes") +
theme(
#  legend.position = "none" #This is to exclude the color legend
) +
  xlim(-5, 5) +
  coord_flip()

```

### enrichGO GO term functions

```{r enrichGO set-up code, eval=FALSE, include=FALSE}
toptable_list <- readRDS("data/toptable_results.RDS")
comparison_name_list <- list("VEH: Human vs Chimp", "IFNy: Human vs Chimp", "Human: IFNy vs VEH", "Chimp: IFNy vs VEH", "Interaction")
enrichGO_setup <- function(toptable, comparison_name) {

  enrichGO_bg <- toptable %>%
    tibble::rownames_to_column("Ensembl") %>%
    pull(Ensembl)

  enrichGO_sig <- toptable %>%
    tibble::rownames_to_column("Ensembl") %>%
    filter(adj.P.Val < 0.05) %>%
    pull(Ensembl)

  if (length(enrichGO_sig) == 0) {
    message("No significant genes for ", comparison_name)
    return(NULL)
  }

  enrichGO_sig_entrez <- bitr(
    enrichGO_sig,
    fromType = "ENSEMBL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  )

  enrichGO_bg_entrez <- bitr(
    enrichGO_bg,
    fromType = "ENSEMBL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db
  )

  list(
      comparison = comparison_name,
      enrichGO_sig_entrez = enrichGO_sig_entrez,
      enrichGO_bg_entrez = enrichGO_bg_entrez
  )
}

enrichGO_results <- vector("list", length(toptable_list))

for (i in seq_along(toptable_list)) {
  enrichGO_results[[i]] <- enrichGO_setup(
    toptable_list[[i]],
    comparison_name_list[[i]]
  )
}

```

```{r Function for clusterProfiler Human analyses, fig.height=6, fig.width=7, eval=FALSE, include=FALSE}
# toptable_list <- readRDS("data/toptable_results.RDS")
# comparison_name_list <- list("VEH: Human vs Chimp", "IFNy: Human vs Chimp", "Human: IFNy vs VEH", "Chimp: IFNy vs VEH", "Interaction")
ontology_list <- list("BP", "MF", "CC")

# bg <- toptable %>%
#     tibble::rownames_to_column("Ensembl") %>%
#     pull(Ensembl)
# 
#   sig <- toptable %>%
#     tibble::rownames_to_column("Ensembl") %>%
#     filter(adj.P.Val < 0.05) %>%
#     pull(Ensembl)
# 
#   if (length(sig) == 0) {
#     message("No significant genes for ", comparison_name)
#     return(NULL)
#   }
# 
#   sig_entrez <- bitr(
#     sig,
#     fromType = "ENSEMBL",
#     toType = "ENTREZID",
#     OrgDb = org.Hs.eg.db
#   )
# 
#   bg_entrez <- bitr(
#     bg,
#     fromType = "ENSEMBL",
#     toType = "ENTREZID",
#     OrgDb = org.Hs.eg.db
#   )
generate_GO_plots <- function(enrichGO_data, ontology){

  ego <- enrichGO(
    gene = enrichGO_data$enrichGO_sig_entrez$ENTREZID,
    universe = enrichGO_data$enrichGO_bg_entrez$ENTREZID,
    OrgDb = org.Hs.eg.db,
    keyType = "ENTREZID",
    ont = ontology,
    pAdjustMethod = "BH",
    readable = TRUE
  )

  if(is.null(ego) || nrow(as.data.frame(ego)) == 0){
    message("No enriched ", ontology,
            " terms for ", enrichGO_data$comparison)
    return(NULL)
  }

  p <- barplot(
    ego,
    showCategory = 20,
    font.size = 10,
    title = paste0(enrichGO_data$comparison, " (GO:", ontology, ")")
  )

  return(p)
}

for(i in seq_along(toptable_list)){

  enrichGO_data <- enrichGO_setup(
    toptable_list[[i]],
    comparison_name_list[[i]]
  )

  if(is.null(enrichGO_data))
    next

  for(ontology in ontology_list){

    p <- generate_GO_plots(enrichGO_data,
                           ontology)

    if(!is.null(p))
      print(p)
  }
}

# human_bg <- V.Human_IFNy_vs_VEH %>%
#   tibble::rownames_to_column("Ensembl") %>%
#   pull(Ensembl)
# 
# human_sig <- human_deg %>%
#   filter(adj.P.Val < 0.05) %>%
#   pull(Ensembl)
# 
# human_sig_entrez <- bitr(human_sig,
#                         fromType = "ENSEMBL",
#                         toType = "ENTREZID",
#                         OrgDb = org.Hs.eg.db)
# 
# human_bg_entrez <- bitr(human_bg,
#                        fromType = "ENSEMBL",
#                        toType = "ENTREZID",
#                        OrgDb = org.Hs.eg.db)
# ego_human <- enrichGO(
#   gene = human_sig_entrez$ENTREZID,
#   universe = human_bg_entrez$ENTREZID,
#   OrgDb = org.Hs.eg.db,
#   keyType = "ENTREZID",
#   ont = "BP",
#   pAdjustMethod = "BH",
#   readable = TRUE
# )
# dotplot(ego_human)

```

### gProfiler2 GO/KEGG term functions

```{r GO analyses setup code, fig.width=5, fig.height=5}
## Setup function
gprofiler2_setup <- function(toptable, comparison_name){

  gprofiler2_bg <- toptable %>%
    tibble::rownames_to_column("Ensembl") %>%
    pull(Ensembl)

  gprofiler2_sig <- toptable %>%
    tibble::rownames_to_column("Ensembl") %>%
    filter(adj.P.Val < 0.05) %>%
    pull(Ensembl)

  if(length(gprofiler2_sig) == 0){

    message("No significant genes for ", comparison_name)

    return(NULL)

  }

  list(
    comparison = comparison_name,
    gprofiler2_sig = gprofiler2_sig,
    gprofiler2_bg = gprofiler2_bg
  )
}

## Creating my input list
gprofiler2_results <- lapply(
  seq_along(toptable_list),
  function(i)
    gprofiler2_setup(
      toptable_list[[i]],
      comparison_name_list[[i]]
    )
)

```

```{r Generating figures illustrating GO/KEGG terms, fig.width=7, fig.height=6}
## Generating a list of databases to query
source_list <- c(
  "GO:BP",
  "GO:MF",
  "GO:CC",
  "KEGG"
)

## Creating the enrichment function
generate_enrichment_results <- function(gprofiler2_data,
                                        source){
  gost_res <- gost(
    query = gprofiler2_data$gprofiler2_sig,
    custom_bg = gprofiler2_data$gprofiler2_bg,
    organism = "hsapiens",
    sources = source,
    correction_method = "fdr"
  )

  if(is.null(gost_res$result) ||
     nrow(gost_res$result) == 0){
    message(
      "No enriched ",
      source,
      " terms for ",
      gprofiler2_data$comparison
    )
    return(NULL)
  }

  gost_res$result %>%
    mutate(
      comparison = gprofiler2_data$comparison,
      source = source
    )
}

## Running the enrichment function
gprofiler2_results_df <- tibble()

for(data in gprofiler2_results){
  if(is.null(data))
    next

  for(source in source_list){
    result <- generate_enrichment_results(
      data,
      source
    )
    if(!is.null(result))
      gprofiler2_results_df <-
        bind_rows(
          gprofiler2_results_df,
          result
        )
  }
}

## Now to plot GO and KEGG Terms separately
for(comp in unique(gprofiler2_results_df$comparison)){
  for(src in unique(gprofiler2_results_df$source)){
    
    plot_df <- gprofiler2_results_df %>%
      filter(
        comparison == comp,
        source == src
      ) %>%
      slice_min(
        order_by = p_value,
        n = 20,
        with_ties = FALSE
      )
    if(nrow(plot_df) == 0)
      next

### Code for dot plot    
    # plot_df <- plot_df %>%
    #   arrange(p_value) %>%
    #   mutate(
    #     term_label = stringr::str_wrap(term_name, width = 40),
    #     term_label = factor(term_label, levels = rev(term_label))
    #     )  
    
    p <- ggplot(
        plot_df,
        aes(
          x = -log10(p_value),
          y = reorder(term_name, -log10(p_value)),
          size = intersection_size)) +
        geom_point(
          color = ifelse(src == "KEGG",
                         "#2E8B57",
                         "#5B4FA2")
        ) +
        
        labs(
          title = paste(comp, "-", src),
          x = "-log10(adjusted p-value)",
          y = NULL,
          size = "GeneCount"
            ) +
        theme_classic() 
      print(p)
  }
}

### Code for bar graph    
#     p <- ggplot(
#       plot_df,
#       aes(
#         x = reorder(term_name, -log10(p_value)),
#         y = -log10(p_value)
#       )) +
#       geom_col(
#         fill = ifelse(src == "KEGG",
#                       "#2E8B57",
#                       "#5B4FA2")
#       ) +
#       coord_flip() +
#       labs(
#         title = paste(comp, "-", src),
#         x = NULL,
#         y = "-log10(Adjusted p-value)"
#       ) +
#       theme_classic()
#     print(p)
#   }
# }

# # The code below is to facet plot all ontology and toptable entities onto one page, not ideal for looking at
# gprofiler2_GO_results_df %>%
#   group_by(comparison, ontology) %>%
#   slice_min(order_by = p_value,
#             n = 20,
#             with_ties = FALSE) %>%
#   ggplot(
#     aes(
#       x = reorder(term_name, -log10(p_value)),
#       y = -log10(p_value)
#     )
#   ) +
#   geom_col(fill = "#5B4FA2") +
#   coord_flip() +
#   facet_grid(
#     ontology ~ comparison,
#     scales = "free_y"
#   ) +
#   labs(
#     x = NULL,
#     y = "-log10(Adj.P-Val)",
#     title = "GO enrichment"
#   ) +
#   theme_bw()
```

### Heatmap comparing -log10(p-value) correlations between species
```{r Comparison Heatmap, fig.width=8, fig.height=5}

human_go <- gost_human$result %>%
  arrange(p_value) %>%
  slice_head(n = 20)

chimp_go <- gost_chimp$result %>%
  arrange(p_value) %>%
  slice_head(n = 20)

top_term_set <- union(
    human_go$term_name,
    chimp_go$term_name
)

human_plot <- human_go %>%
    select(term_name, p_value) %>%
    mutate(Human = -log10(p_value))

chimp_plot <- chimp_go %>%
    select(term_name, p_value) %>%
    mutate(Chimp = -log10(p_value))

heatmap_df_1 <- full_join(
    human_plot,
    chimp_plot,
    by = "term_name"
)

heatmap_df_2 <- heatmap_df_1 %>%
    mutate(
        Human = replace_na(Human, 0),
        Chimp = replace_na(Chimp, 0)
    )

heatmap_matrix <- heatmap_df_2 %>%
  dplyr::select(
    term_name,
    Human,
    Chimp
  ) %>% 
  column_to_rownames("term_name") %>%
  as.matrix()

#heatmap_matrix

Heatmap(
  heatmap_matrix,
  name = "-log10(adj.P)",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  row_names_side = "right",
  row_names_gp = gpar(fontsize = 8),
  row_names_max_width = unit(12, "cm")
  )

```


### Previous single-pass code for enrichGO
```{r GO CC analyses, fig.height=5, fig.width=5, eval=FALSE, include=FALSE}
# generate_GO_plots <- function(toptable, comparison_name){
# 
#   bg <- toptable %>%
#     tibble::rownames_to_column("Ensembl") %>%
#     pull(Ensembl)
# 
#   sig <- toptable %>%
#     tibble::rownames_to_column("Ensembl") %>%
#     filter(adj.P.Val < 0.05) %>%
#     pull(Ensembl)
# 
#   if(length(sig) == 0){
#     message("No significant genes for ", comparison_name)
#     return(NULL)
#   }
# 
#   sig_entrez <- bitr(
#     sig,
#     fromType = "ENSEMBL",
#     toType = "ENTREZID",
#     OrgDb = org.Hs.eg.db
#   )
# 
#   bg_entrez <- bitr(
#     bg,
#     fromType = "ENSEMBL",
#     toType = "ENTREZID",
#     OrgDb = org.Hs.eg.db
#   )
# 
#   ego <- enrichGO(
#     gene = sig_entrez$ENTREZID,
#     universe = bg_entrez$ENTREZID,
#     OrgDb = org.Hs.eg.db,
#     keyType = "ENTREZID",
#     ont = "CC", # or other ont classifications such as "BP" or "MF"
#     pAdjustMethod = "BH",
#     readable = TRUE
#   )
# 
#   if (is.null(ego) || nrow(as.data.frame(ego)) == 0) {
#   message("No enriched GO terms for ", comparison_name)
#   return(NULL)
# }
#   
#   p <- barplot(
#     ego,
#     showCategory = 20,
#     font.size = 10,
#     title = paste0("GO_CC ", comparison_name) # replcae CC with BP or MF as needed
#   )
# 
#   return(p)
# }
# 
# for(i in seq_along(save_toptable_list)){
# 
#   p <- generate_GO_plots(
#     save_toptable_list[[i]],
#     comparison_name_list[[i]]
#   )
# 
#   if(!is.null(p))
#     print(p)
# }


```

## GSEA of IFNy enriched responses

```{r GSEA of Human vs Chimp IFNy, fig.width=6, fig.height=10}

# Human ranked gene list
human_results <- V.Human_IFNy_vs_VEH %>%
  tibble::rownames_to_column("Ensemble") %>%
  left_join(
    counts %>%
      dplyr::select(Ensemble, Entrez_ID, Symbol),
    by = "Ensemble"
  ) %>%
  dplyr::filter(!is.na(Entrez_ID)) %>%
  dplyr::distinct(Entrez_ID, .keep_all = TRUE)

human_rank <- human_results$logFC
names(human_rank) <- human_results$Entrez_ID
human_rank <- sort(human_rank, decreasing = TRUE)


# Chimp ranked gene list
chimp_results <- V.Chimp_IFNy_vs_VEH %>%
  tibble::rownames_to_column("Ensemble") %>%
  left_join(
    counts %>%
      dplyr::select(Ensemble, Entrez_ID, Symbol),
    by = "Ensemble"
  ) %>%
  dplyr::filter(!is.na(Entrez_ID)) %>%
  dplyr::distinct(Entrez_ID, .keep_all = TRUE)

chimp_rank <- chimp_results$logFC
names(chimp_rank) <- chimp_results$Entrez_ID
chimp_rank <- sort(chimp_rank, decreasing = TRUE)


# Run GSEA
human_gsea <- gseGO(
    geneList = human_rank,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    keyType = "ENTREZID",
    pAdjustMethod = "BH"
)

chimp_gsea <- gseGO(
    geneList = chimp_rank,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    keyType = "ENTREZID",
    pAdjustMethod = "BH"
)


# Extract results
human_gsea_results <- human_gsea@result
chimp_gsea_results <- chimp_gsea@result

# Removing duplicate comparison IDs
human_gsea_results <- human_gsea@result %>%
  dplyr::filter(!is.na(ID))

chimp_gsea_results <- chimp_gsea@result %>%
  dplyr::filter(!is.na(ID))

# Compare pathways
comparison <- inner_join(
  human_gsea_results,
  chimp_gsea_results,
  by = "ID",
  suffix = c("_Human", "_Chimp")
)


# Pathways to label
label_df <- comparison %>%
  filter(
    p.adjust_Human < 0.05 |
    p.adjust_Chimp < 0.05
  )


# # Scatter plot
# ggplot(
#   comparison,
#   aes(
#     x = NES_Chimp,
#     y = NES_Human
#   )
# ) +
# 
#   geom_point(
#     color = "steelblue",
#     alpha = 0.7,
#     size = 2
#   ) +
# 
#   geom_abline(
#     slope = 1,
#     intercept = 0,
#     linetype = "dashed",
#     color = "red"
#   ) +
# 
#   geom_text_repel(
#     data = label_df,
#     aes(label = Description_Human),
#     size = 3,
#     max.overlaps = 30,
#     box.padding = 0.5,
#     point.padding = 0.2
#   ) +
# 
#   coord_equal() +
# 
#   theme_classic(base_size = 14) +
# 
#   labs(
#     title = "GSEA comparison: Human vs Chimp",
#     x = "Chimp NES",
#     y = "Human NES"
#   )

# Heatmap for all pathways
heatmap_matrix <- comparison %>%
  dplyr::select(
    ID,
    NES_Human,
    NES_Chimp
  ) %>%
  tibble::column_to_rownames("ID") %>%
  as.matrix()

heatmap_matrix[is.na(heatmap_matrix)] <- 0

Heatmap(
  heatmap_matrix,
  name = "NES",
  col = colorRamp2(
    c(-3, 0, 3),
    c("blue", "white", "red")
  ),
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  row_names_gp = gpar(fontsize = 8),
  column_title = "GSEA Comparison",
  column_names_gp = gpar(fontsize = 12)
)

# Heatmap for significant pathways
comparison_sig <- comparison %>%
  filter(
    p.adjust_Human < 0.05 |
    p.adjust_Chimp < 0.05
  )

heatmap_matrix <- comparison_sig %>%
    select(
        ID,
        NES_Human,
        NES_Chimp
    ) %>%
    column_to_rownames("ID") %>%
    as.matrix()

heatmap_matrix[is.na(heatmap_matrix)] <- 0

Heatmap(
  heatmap_matrix,
  name = "NES",
  col = colorRamp2(
    c(-3, 0, 3),
    c("blue", "white", "red")
  ),
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  row_names_gp = gpar(fontsize = 8),
  column_title = "GSEA Comparison",
  column_names_gp = gpar(fontsize = 12)
)

# Heatmap for top/bottom 'n' significant pathways
top_pathways <- comparison %>%
    arrange(p.adjust_Human + p.adjust_Chimp) %>%
    slice_head(n = 20)

heatmap_matrix <- top_pathways %>%
    select(
        ID,
        NES_Human,
        NES_Chimp
    ) %>%
    column_to_rownames("ID") %>%
    as.matrix()

heatmap_matrix[is.na(heatmap_matrix)] <- 0

Heatmap(
    heatmap_matrix,
    name = "NES",
    col = colorRamp2(
        c(-3,0,3),
        c("blue","white","red")
    ),
    cluster_rows = TRUE,
    cluster_columns = FALSE
)
```

